Exploring probability distributions for bivariate temporal granularities

### 2019-08-12 main.Rmd
spell check and making implememtnation and concept different for multiple gran construction

<!--temporal granularities and why should we care --> 
Temporal data can be available at various resolution depending on the context. Social and economic data are often collected and reported at coarses temporal scales like monthly, quarterly or annually. But with recent advancement in technology, more and more data are recorded and stored at much finer temporal scales than that was previously possible. It might be sufficient to observe energy consumption every half an hour, but energy supply needs to be monitored every minute and number of web searches requires optimisation every second.  As the frequency of data increases, the number of questions about the observed variable that need to be addressed also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales like days, months or quarters. This approach requires deconstructing time in various possible ways.

@@

The motivation for this work comes from the desire to provide methods to better understand large quantities of measurements on energy usage reported by smart meters in household across Australia, and indeed many parts of the world. Smart meters currently provide half-hourly use in kwh for each household, from the time that they were installed, some as early as 2012. Households are distributed geographically, and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks.

It is common to see aggregates of usage across households, total kwh used each half hour by state, for example, because energy companies need to understand maximum loads that they will have to plan ahead to accommodate. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency.

@@

<!--existing closely related work and how ours relate --> 
 This paper utilises the nestedness of time granularities to obtain multiple-order-up granularities from single-order-up ones.

<!--actions on multiple order up - usage --> 
Finally, visualizing data across single/multiple order-up granularities help us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. However, this approach can lead to a myriad of choices all of which are not useful. Analysts are expected to iteratively visualise these choices for exploring possible patterns in the data. But too many choices might leave him bewildered.

@@

# Formal conceptualisation of calendar categorisations{#cirgran-def}

<!-- general calendar categorisation -->
Often we partition time into months, weeks, days and so on in a hierarchical manner to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Examples of time abstractions may also include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, workingday/non-workingday, etc which are useful to represent different periodicities in the data. Let us call all of these different abstractions of time as "calendar categorizations".

These calendar categorisations can be linear, circular or aperiodic. The calendar categorisations as "linear" if they respect the linear progression of time. We call these **linear time granularities**. Examples include hours, days, weeks and months. **Circular time granularities** accommodate periodicities in time such as hour of the day, and day of the week. **Aperiodic time granularities** are neither linear nor circular, such as day of the month or public holidays.

@@

<!-- <!-- What is calendar categorisation --> 
<!-- What we need to add to this are additional categorizations of time that are not linear granularities and are useful to represent periodicity. Examples include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, workingday/non-workingday, etc. Many of these are circular, such as day-of-week, time-of-day. Some are nearly circular such as day-of-month. Some are irregular such as workingday/non-working day. Let's call all of these "calendar categorizations". Anything that maps a time index to a categorical variable can be considered a **calendar categorization**.  -->

@@

**Defintion:** An **Aperiodic circular granularity** can not be defined using modular arithmetic in a similar fashion. The modulus for these type of calendar categorisations are not constant due to unequal length of some linear granularities. For example, please refer to the table below:

@@

 <!-- Why we need single order up and multiple order up granularities -->
<!-- A time series which is characterised by being composed of repeating cycles. -->

Identifying repeating (periodic/aperiodic) patterns are necessary in revealing patterns and future trends of a temporal data. Often there is a need for periodicity detection to find whether and how frequent a periodic/aperiodic pattern is repeated within the series. To consider the exhaustive set of temporal regularities that might exist ein the data, we can categorise the set of temporal granularities to single or multiple order up granularities.

@@

The index of a tsibble can be any date-time object which comes in different formats. [@Grolemund2011-qf] goes a long way in facilitating analysis of date-time data and extracting each date-time element with its own accessor function. However, table shows the accessor functions mostly allows for the creation of single-order-up granularities.

```{r table_accessor}
kable(tibble::tibble(
  `Categorisation` = c("Second-of-minute", "Minute-of-hour", "Hour-of-day", "Day-of-week", "Day-of-month", "Week-of-year", "Day-of-year","Month-of-year"),
 Type = c("Periodic", "Periodic", "Periodic", "Periodic", "Aperiodic", "Aperiodic", "Aperiodic", "Periodic"),
Accessor = c("second()", "minute()", "hour()","wday()", "mday()","week()", "yday()", "month()")), format = "markdown")
```

The hierarchical nature of time helps to create a framework where the single-order-up granularities can be used recursively to create multiple-order-up granularities.The idea is to incorporate and access different calendar categorisations to account for varied levels of periodicities in the data. The computation of multiple-order-up granularities from single-order-up granularities can differ basis the single-order-up granularities are circular or aperiodic.

@@

There can be three scenarios for obtaining calendar categorisation here:
- calendar categorisation consisting of two linear granularities whose orders are less than day  
- calendar categorisation consisting of two linear granularities whose orders are more than month  
- calendar categorisation consisting of one linear granularity with order at most day and another with order at least month

The calendar categorisation resulting from the first two cases are periodic and has been handled in the earlier section.

The calendar categorisation resulting from the last case are aperiodic. Examples might include day of the quarter or hour of the month. In this section, we will see how to obtain aperiodic circular granularities of these types.

@@

VIC%>% filter(year %in% c(2012, 2013, 2014),Indx_Day_Month %in% c(1,15,29,31))%>% ggplot(aes(as.factor(yday),OperationalLessIndustrial,group = yday)) + geom_boxplot()+ facet_wrap(~Indx_Day_Month) + ylab("Electricity Demand [KWh]") +
  xlab("Days of the Year") + scale_x_discrete(breaks=seq(0,366,60)) +theme(legend.position = "bottom",strip.text = element_text(size = 7, margin = margin())) + ggtitle("(c) Boxplot")

@@

<!-- We investigate some combinations of circular time granularities which facilitate or hinder exploratory analysis. Since we are restricting ourselves to 2D space, we assume that the circular granularities are plotted across the aesthetics and facets. The combinations of circular granularities which promote the exploratory analysis through visualization are referred to as **harmonies** and the ones which impede the analysis are referred to as **clashes**. -->

>>>>

<!--temporal granularities and why should we care --> 
Temporal data can be available at various resolution depending on the context. Social and economic data are often collected and reported at coarser temporal scales like monthly, quarterly or annually. But with recent advancement in technology, more and more data are recorded and stored at much finer temporal scales than that was previously possible. It might be sufficient to observe energy consumption every half an hour, but energy supply needs to be monitored every minute and number of web searches requires optimization every second.  As the frequency of data increases, the number of questions about the observed variable that need to be addressed also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales like days, months or quarters. This approach requires deconstructing time in various possible ways.

@@

The motivation for this work comes from the desire to provide methods to better understand large quantities of measurements on energy usage reported by smart meters in household across Australia, and indeed many parts of the world. Smart meters currently provide half-hourly use in kWh for each household, from the time that they were installed, some as early as 2012. Households are distributed geographically, and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks.

It is common to see aggregates of usage across households, total kWh used each half hour by state, for example, because energy companies need to understand maximum loads that they will have to plan ahead to accommodate. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency.

@@

<!--existing closely related work and how ours relate --> 
 This paper utilizes the nestedness of time granularities to obtain multiple-order-up granularities from single-order-up ones.

<!--actions on multiple order up - usage --> 
Finally, visualizing data across single/multiple order-up granularities help us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. However, this approach can lead to a myriad of choices all of which are not useful. Analysts are expected to iteratively visualize these choices for exploring possible patterns in the data. But too many choices might leave him bewildered.

@@

# Formal conceptualization of calendar categorizations{#cirgran-def}

<!-- general calendar categorization -->
Often we partition time into months, weeks, days and so on in a hierarchical manner to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Examples of time abstractions may also include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc which are useful to represent different periodicities in the data. Let us call all of these different abstractions of time as "calendar categorizations".

These calendar categorizations can be linear, circular or aperiodic. The calendar categorizations as "linear" if they respect the linear progression of time. We call these **linear time granularities**. Examples include hours, days, weeks and months. **Circular time granularities** accommodate periodicities in time such as hour of the day, and day of the week. **Aperiodic time granularities** are neither linear nor circular, such as day of the month or public holidays.

@@

<!-- <!-- What is calendar categorization --> 
<!-- What we need to add to this are additional categorizations of time that are not linear granularities and are useful to represent periodicity. Examples include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc. Many of these are circular, such as day-of-week, time-of-day. Some are nearly circular such as day-of-month. Some are irregular such as working day/non-working day. Let's call all of these "calendar categorizations". Anything that maps a time index to a categorical variable can be considered a **calendar categorization**.  -->

@@

**Definition:** An **Aperiodic circular granularity** can not be defined using modular arithmetic in a similar fashion. The modulus for these type of calendar categorizations are not constant due to unequal length of some linear granularities. For example, please refer to the table below:

@@

 <!-- Why we need single order up and multiple order up granularities -->
<!-- A time series which is characterized by being composed of repeating cycles. -->

Identifying repeating (periodic/aperiodic) patterns are necessary in revealing patterns and future trends of a temporal data. Often there is a need for periodicity detection to find whether and how frequent a periodic/aperiodic pattern is repeated within the series. To consider the exhaustive set of temporal regularities that might exist in the data, we can categorize the set of temporal granularities to single or multiple order up granularities.

@@

<!-- The index of a tsibble can be any date-time object which comes in different formats. [@Grolemund2011-qf] goes a long way in facilitating analysis of date-time data and extracting each date-time element with its own accessor function. However, table shows the accessor functions mostly allows for the creation of single-order-up granularities.  -->

<!-- ```{r table_accessor} -->
<!-- kable(tibble::tibble( -->
<!--   `Categorization` = c("Second-of-minute", "Minute-of-hour", "Hour-of-day", "Day-of-week", "Day-of-month", "Week-of-year", "Day-of-year","Month-of-year"), -->
<!--  Type = c("Periodic", "Periodic", "Periodic", "Periodic", "Aperiodic", "Aperiodic", "Aperiodic", "Periodic"), -->
<!-- Accessor = c("second()", "minute()", "hour()","wday()", "mday()","week()", "yday()", "month()")), format = "markdown") -->
<!-- ``` -->

The hierarchical nature of time helps to create a framework where the single-order-up granularities can be used recursively to create multiple-order-up granularities.The idea is to incorporate and access all possible calendar categorizations to account for different periodicities in the data. The computation of multiple-order-up granularities from single-order-up granularities can differ basis the single-order-up granularities are circular or aperiodic.

@@

There can be three scenarios for obtaining calendar categorization here:
- calendar categorization consisting of two linear granularities whose orders are less than day  
- calendar categorization consisting of two linear granularities whose orders are more than month  
- calendar categorization consisting of one linear granularity with order at most day and another with order at least month

The calendar categorization resulting from the first two cases are periodic and has been handled in the earlier section.

The calendar categorization resulting from the last case are aperiodic. Examples might include day of the quarter or hour of the month. In this section, we will see how to obtain aperiodic circular granularities of these types.

@@

VIC%>% filter(year %in% c(2012, 2013, 2014),Indx_Day_Month %in% c(1,15,29,31))%>% ggplot(aes(as.factor(yday),OperationalLessIndustrial,group = yday)) + geom_boxplot()+ facet_wrap(~Indx_Day_Month) + ylab("Electricity Demand [KWh]") +
  xlab("Days of the Year") + scale_x_discrete(breaks=seq(0,366,60)) +theme(legend.position = "bottom",strip.text = element_text(size = 7, margin = margin())) + ggtitle("(c) Box plot")

@@

Levels are categorized as very high/high/medium/low each for the facet variable and the x-axis variable. Default values for these levels are based on levels of common temporal granularities like day of the month, day of a fortnight or day of a week. For example, any levels above 31 is considered as very high, any levels between 14 to 31 are taken as high and that between 7 to 14 is taken as medium and below 7 is low. 31, 14 and 7 are the levels of days-of-month, days-of-fortnight and days-of week respectively. These default values are decided based on usual cognitive power while comparing across facets and display size available to us. Let us consider case by case and see which plots are better suitable in which scenarios.

 -  very high facet and x-axis levels

 X-axis variable is treated as a categorical variable and hence any plots which
